### 2025-08-17 chapters/benchmark.tex
4.4 complete

% ------------------------
% Wall-clock time figures
% ------------------------
Figure~\ref{fig:wallclock_coremark} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes. Execution time remained consistent, with the sequential case averaging approximately 101.17 seconds of wall-clock time and the parallel case averaging around 103.37 seconds. This corresponds to an effective throughput of about 988.4 iterations per second for the sequential run and 967.4 iterations per second for the parallel run. User-mode CPU time dominated in both cases (101.20 seconds sequential, 103.39 seconds parallel), accounting for over 99.9\% of total CPU time. System-mode CPU time was negligible (under 0.03 seconds), confirming that the workload’s operational footprint is almost entirely in user space with minimal reliance on system calls or kernel transitions.

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/wallclock_coremark.png}
%   \caption{CoreMark wall-clock time in native sequential and parallel runs.}
%   \label{fig:wallclock_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/wallclock_dhrystone.png}
%   \caption{Dhrystone wall-clock time in native sequential and parallel runs.}
%   \label{fig:wallclock_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/wallclock_kyber.png}
%   \caption{Kyber wall-clock time in native sequential and parallel runs.}
%   \label{fig:wallclock_kyber}
% \end{figure}

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/cputime_coremark.png}
%   \caption{CoreMark CPU time (user and system) in native sequential and parallel runs.}
%   \label{fig:cputime_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/cputime_dhrystone.png}
%   \caption{Dhrystone CPU time (user and system) in native sequential and parallel runs.}
%   \label{fig:cputime_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/cputime_kyber.png}
%   \caption{Kyber CPU time (user and system) in native sequential and parallel runs.}
%   \label{fig:cputime_kyber}
% \end{figure}

% ------------------------
% Throughput figures
% ------------------------
Figure~\ref{fig:dhrystone_native} presents detailed results from the Dhrystone benchmark under native conditions. The sequential runs exhibited an average elapsed time of approximately 101.2 seconds per run, with user-mode CPU time closely matching this figure, indicating near-total CPU utilization. In contrast, the parallel merged runs completed in roughly 103.4 seconds, with user CPU time again nearly equal to wall-clock time. Throughput, measured in Dhrystone operations per second (DPS), was higher in the sequential case—approximately 988.4~DPS versus 967.4~DPS in the parallel configuration—reflecting slight scheduling overhead from virtual concurrency under QEMU.

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/throughput_coremark.png}
%   \caption{CoreMark throughput in native sequential and parallel runs.}
%   \label{fig:throughput_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/throughput_dhrystone.png}
%   \caption{Dhrystone throughput in native sequential and parallel runs.}
%   \label{fig:throughput_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/throughput_kyber.png}
%   \caption{Kyber throughput in native sequential and parallel runs.}
%   \label{fig:throughput_kyber}
% \end{figure}

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_coremark.png}
%   \caption{CoreMark resident set size (RSS) in native sequential and parallel runs.}
%   \label{fig:memory_rss_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_dhrystone.png}
%   \caption{Dhrystone resident set size (RSS) in native sequential and parallel runs.}
%   \label{fig:memory_rss_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/memory_rss_kyber.png}
%   \caption{Kyber resident set size (RSS) in native sequential and parallel runs.}
%   \label{fig:memory_rss_kyber}
% \end{figure}

% ------------------------
% Page faults figures
% ------------------------
% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_coremark.png}
%   \caption{CoreMark minor and major page faults in native sequential and parallel runs.}
%   \label{fig:pagefaults_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_dhrystone.png}
%   \caption{Dhrystone minor and major page faults in native sequential and parallel runs.}
%   \label{fig:pagefaults_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/pagefaults_kyber.png}
%   \caption{Kyber minor and major page faults in native sequential and parallel runs.}
%   \label{fig:pagefaults_kyber}
% \end{figure}

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_coremark.png}
%   \caption{CoreMark voluntary and involuntary context switches in native sequential and parallel runs.}
%   \label{fig:contextswitches_coremark}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_dhrystone.png}
%   \caption{Dhrystone voluntary and involuntary context switches in native sequential and parallel runs.}
%   \label{fig:contextswitches_dhrystone}
% \end{figure}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/contextswitches_kyber.png}
%   \caption{Kyber voluntary and involuntary context switches in native sequential and parallel runs.}
%   \label{fig:contextswitches_kyber}
% \end{figure}

% ------------------------
% Cycle counts figures (Kyber only)
% ------------------------
Kyber performance was evaluated using the official \texttt{test\_speed512} binary from the reference implementation's \texttt{ref/} directory.\footnote{See \url{https://github.com/pq-crystals/kyber}. The \texttt{test\_speed512} utility executes keypair generation, encapsulation, and decapsulation 1000 times each and reports cycle counts using architecture-level counters emulated under QEMU.} Table~\ref{tab:kyber_run10_cycles} summarizes the median cycle counts for each operation, while Figure~\ref{fig:kyber_native_cycles} compares sequential and parallel execution.

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.9\linewidth]{figures/cyclecounts_kyber.png}
%   \caption{Kyber median cycle counts for keypair generation, encapsulation, and decapsulation in native sequential and parallel runs.}
%   \label{fig:kyber_native_cycles}
% \end{figure}

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
    \caption{Comparison of Kyber Median CPU Cycles per Operation: Sequential Single-Threaded (\texttt{run10\_SEQ}) vs Parallel Merged (\texttt{run10}) Execution (10 Runs)}
    \label{fig:kyber_native_cycles}
\end{figure}
All cryptographic operations executed entirely in user space, without invoking privileged instructions or incurring system-level overhead. Memory footprint remained minimal (peak RSS ~10~KB), with only a single minor page fault observed across all runs.

\begin{table}[H]
\centering
\scriptsize
\caption{Median CPU Cycles for Kyber Operations}
\begin{tabular}{|l|r|r|}
\hline
\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} \\
\hline
indcpa\_keypair & 1,870,200 & 1,702,800 \\
\hline
indcpa\_enc     & 38,376,800 & 2,174,400 \\
\hline
indcpa\_dec     & 39,060,800 & 618,400 \\
\hline
kyber\_keypair  & 1,870,200 & 1,800,300 \\
\hline
kyber\_encaps   & 38,376,800 & 2,285,900 \\
\hline
kyber\_decaps   & 39,060,800 & 2,921,800 \\
\hline
\end{tabular}
\label{tab:kyber_median_cycles}
\end{table}

@@

All benchmarks were initially executed ten times sequentially under enclave protection to establish a direct, low-variability comparison against native runs. The Kyber \texttt{kyber\_decaps} operation exhibited median CPU cycle counts consistently around 3.5 million cycles per run (e.g., 3,505,686 cycles in Run~1), approximately 1.2$\times$ the native sequential baseline of about 2.9 million cycles (see Table~\ref{tab:kyber_median_cycles}). Average cycles were slightly higher, around 3.53 million, reflecting stable and repeatable enclave performance. Other key Kyber operations, such as \texttt{indcpa\_keypair} and \texttt{indcpa\_enc}, showed median cycle counts near 1.74 million and 2.62 million cycles respectively, consistent across runs.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/kyber_64M_1Core_comparison.pdf}
    \caption{Cycle count comparison for Kyber operations (native vs enclave). Error bars reflect variation over 10 runs.}
    \label{fig:kyber_cycles}
\end{figure}

@@

%\begin{figure}[h]
%    \centering
%    \includegraphics[width=0.7\textwidth]{figures/wall_clock_comparison.pdf}
%    \caption{Wall-clock execution time across benchmarks (native vs enclave, averaged over 10 runs).}
%    \label{fig:wall_clock}
%\end{figure}

@@

@@

@@

These findings confirm that secure enclaves can sustain a wide range of computational workloads with predictable overheads and stable memory behavior, while also revealing important hardware-imposed concurrency limitations. The detailed performance and system metric breakdown presented here forms a critical foundation for future optimization and hardware enhancement efforts targeting secure enclave scalability and efficiency.

\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}

The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications. Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.

Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution. This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.

As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design. Memory usage during enclave execution remained stable, with no indications of paging or contention. System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.

From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed. This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.

\section{Impact of Hardware Parameters (Cores, Memory, Cache)}

To comprehensively assess how hardware configurations affect enclave performance, we systematically varied CPU core counts, memory allocations, and cache configurations.

Scaling the number of CPU cores yielded anticipated improvements for parallel workloads, with throughput gains plateauing beyond six cores due to synchronization overhead and increased enclave management complexity. Sequential workloads such as Kyber remained unaffected by core scaling.

Memory allocation experiments revealed that increasing memory size from 64MB to 2GB improved performance for memory-intensive benchmarks by reducing paging and increasing cache hit rates. However, beyond approximately 512MB, gains diminished, indicating sufficient memory provision for the workload’s working set.

Cache configuration impacts were evaluated using the QEMU cache modeling plugin. Larger cache sizes and higher associativity reduced miss rates, particularly for instruction-heavy workloads like Dhrystone. While the plugin’s system-wide scope limits insights into enclave-specific cache activity, the trends suggest that cache tuning can meaningfully influence enclave performance.

Importantly, the relationship between memory allocation and PMP usage is nuanced. Larger enclave memory regions may require multiple PMP entries if the memory cannot be described contiguously in a single PMP register, effectively reducing the total number of enclaves supported simultaneously. This interaction between memory layout and PMP capacity represents a critical consideration for enclave deployment strategies.

>>>>

% ------------------------
% Wall-clock time figures
% ------------------------
Figure~\ref{fig:coremark_perf_comparison_log} reports CoreMark's performance profile using the latest data from Run~10 of both sequential and parallel execution modes. Execution time remained consistent, with the sequential case averaging approximately 101.17 seconds of wall-clock time and the parallel case averaging around 103.37 seconds. This corresponds to an effective throughput of about 988.4 iterations per second for the sequential run and 967.4 iterations per second for the parallel run. User-mode CPU time dominated in both cases (101.20 seconds sequential, 103.39 seconds parallel), accounting for over 99.9\% of total CPU time. System-mode CPU time was negligible (under 0.03 seconds), confirming that the workload’s operational footprint is almost entirely in user space with minimal reliance on system calls or kernel transitions.

\begin{figure}[htbp]
    \centering
    \begin{subfigure}{\textwidth}
        \centering
        \begin{tikzpicture}
        \begin{axis}[
            ybar,
            bar width=8pt,
            width=\textwidth,
            height=8cm,
            ymode=log,
            log origin=infty,
            ymin=0.001,
            ylabel={Value (log scale)},
            symbolic x coords={
                Total\ Time,
                Iterations/Sec,
                User\ CPU\ Time,
                System\ CPU\ Time,
                Max\ RSS,
                Minor\ PF,
                Major\ PF,
                Voluntary\ CS,
                Involuntary\ CS
            },
            xtick=data,
            x tick label style={rotate=30, anchor=east},
            enlarge x limits=0.05,
            ymajorgrids,
            grid style=dashed,
            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1}
        ]
        \addplot[pattern=north east lines,pattern color=black] coordinates {
            (Total\ Time,103.24)
            (Iterations/Sec,968.64)
            (User\ CPU\ Time,103.26)
            (System\ CPU\ Time,0.02)
            (Max\ RSS,1712.00)
            (Minor\ PF,83.60)
            (Major\ PF,0.20)
            (Voluntary\ CS,0.70)
            (Involuntary\ CS,25815.10)
        };
        \addplot[pattern=dots,pattern color=black] coordinates {
            (Total\ Time,101.11)
            (Iterations/Sec,989.00)
            (User\ CPU\ Time,101.14)
            (System\ CPU\ Time,0.01)
            (Max\ RSS,1691.60)
            (Minor\ PF,83.60)
            (Major\ PF,0.20)
            (Voluntary\ CS,0.90)
            (Involuntary\ CS,350.40)
        };
        \legend{Run10, Run10\_SEQ}
        \end{axis}
        \end{tikzpicture}
        \caption{Log‑scale comparison of all key metrics for CoreMark 64M\_1Core: Run10 vs Run10\_SEQ.}
        \label{fig:coremark_perf_comparison_log}
    \end{subfigure}

    \vspace{1.2cm}

    \begin{subfigure}{\textwidth}
        \centering
        \begin{tikzpicture}
        \begin{axis}[
            ybar,
            bar width=8pt,
            width=\textwidth,
            height=8cm,
            ymode=log,
            log origin=infty,
            ymin=0.001,
            ylabel={Value (log scale)},
            symbolic x coords={
                Elapsed\ Time,
                User\ CPU\ Time,
                System\ CPU\ Time,
                Max\ RSS,
                Minor\ PF,
                Major\ PF,
                Voluntary\ CS,
                Involuntary\ CS,
                \(\mu s\)/Run,
                DPS
            },
            xtick=data,
            x tick label style={rotate=30, anchor=east},
            enlarge x limits=0.05,
            ymajorgrids,
            grid style=dashed,
            legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1}
        ]
        \addplot[pattern=north east lines,pattern color=black] coordinates {
            (Elapsed\ Time,103.74)
            (User\ CPU\ Time,11.04)
            (System\ CPU\ Time,0.02)
            (Max\ RSS,1700.00)
            (Minor\ PF,89.60)
            (Major\ PF,0.30)
            (Voluntary\ CS,0.70)
            (Involuntary\ CS,2583.60)
            (\(\mu s\)/Run,10.10)
            (DPS,96478.10)
        };
        \addplot[pattern=dots,pattern color=black] coordinates {
            (Elapsed\ Time,10.02)
            (User\ CPU\ Time,10.03)
            (System\ CPU\ Time,0.01)
            (Max\ RSS,1700.00)
            (Minor\ PF,88.60)
            (Major\ PF,0.10)
            (Voluntary\ CS,0.80)
            (Involuntary\ CS,44.00)
            (\(\mu s\)/Run,0.50)
            (DPS,998646.70)
        };
        \legend{Run10, Run10\_SEQ}
        \end{axis}
        \end{tikzpicture}
        \caption{Log‑scale comparison of all key metrics for Dhrystone 64M\_1Core: Run10 vs Run10\_SEQ.}
        \label{fig:dhrystone_perf_comparison_log}
    \end{subfigure}
    \caption{Comparison of CoreMark and Dhrystone benchmark performance metrics (log-scale) under Run10 and Run10\_SEQ modes.}
\end{figure}

@@

% ------------------------
% Throughput figures
% ------------------------
Figure~\ref{fig:dhrystone_perf_comparison_log} presents detailed results from the Dhrystone benchmark under native conditions. The sequential runs exhibited an average elapsed time of approximately 101.2 seconds per run, with user-mode CPU time closely matching this figure, indicating near-total CPU utilization. In contrast, the parallel merged runs completed in roughly 103.4 seconds, with user CPU time again nearly equal to wall-clock time. Throughput, measured in Dhrystone operations per second (DPS), was higher in the sequential case—approximately 988.4~DPS versus 967.4~DPS in the parallel configuration—reflecting slight scheduling overhead from virtual concurrency under QEMU.

@@

@@

% ------------------------
% Cycle counts figures (Kyber only)
% ------------------------
Kyber performance was evaluated using the official \texttt{test\_speed512} binary from the reference implementation's \texttt{ref/} directory.\footnote{See \url{https://github.com/pq-crystals/kyber}. The \texttt{test\_speed512} tool benchmarks key operations using emulated cycle counters in QEMU.} Figure~\ref{fig:kyber_median_cycles_chart_log} presents the median cycle counts for each cryptographic operation, comparing both sequential and parallel enclave execution.

% \begin{figure}[htbp]
%     \centering
%     \includegraphics[width=0.9\linewidth]{figures/kyber_64M_1Core.pdf}
%     \caption{Comparison of Kyber Median CPU Cycles per Operation: Sequential Single-Threaded (\texttt{run10\_SEQ}) vs Parallel Merged (\texttt{run10}) Execution (10 Runs)}
%     \label{fig:kyber_native_cycles}
% \end{figure}

All cryptographic operations executed entirely in user space, without invoking privileged instructions or incurring system-level overhead. Memory footprint remained minimal (peak RSS ~10~KB), with only a single minor page fault observed across all runs.

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=10pt,
    width=14cm,
    height=8cm,
    ymode=log,
    log origin=infty,
    ymin=1e5,
    ymax=1e9,
    ytick={1e5,1e6,1e7,1e8,1e9},
    yticklabels={10$^5$,10$^6$,10$^7$,10$^8$,10$^9$},
    ylabel={Median CPU Cycles (log scale)},
    symbolic x coords={
        indcpa\_keypair,
        indcpa\_enc,
        indcpa\_dec,
        kyber\_keypair,
        kyber\_encaps,
        kyber\_decaps
    },
    xtick=data,
    x tick label style={rotate=30, anchor=east},
    enlarge x limits=0.05,
    ymajorgrids,
    grid style=dashed,
    legend style={at={(0.5,-0.25)},anchor=north,legend columns=-1},
    nodes near coords,
    every node near coord/.append style={font=\scriptsize, rotate=90, anchor=west}
]
% Parallel values
\addplot[pattern=north east lines,pattern color=black] coordinates {
    (indcpa\_keypair,1870200)
    (indcpa\_enc,38376800)
    (indcpa\_dec,39060800)
    (kyber\_keypair,1870200)
    (kyber\_encaps,38376800)
    (kyber\_decaps,39060800)
};
% Sequential values
\addplot[pattern=dots,pattern color=black] coordinates {
    (indcpa\_keypair,1702800)
    (indcpa\_enc,2174400)
    (indcpa\_dec,618400)
    (kyber\_keypair,1800300)
    (kyber\_encaps,2285900)
    (kyber\_decaps,2921800)
};

\legend{Parallel, Sequential}
\end{axis}
\end{tikzpicture}
\caption{Median CPU cycles for Kyber operations in parallel vs sequential modes, plotted on a logarithmic scale to make smaller values visible alongside large ones.}
\label{fig:kyber_median_cycles_chart_log}
\end{figure}

@@

All benchmarks were initially executed ten times sequentially under enclave protection to establish a direct, low-variability comparison against native runs. The Kyber \texttt{kyber\_decaps} operation exhibited median CPU cycle counts consistently around 3.5 million cycles per run (e.g., 3,505,686 cycles in Run~1), approximately 1.2$\times$ the native sequential baseline of about 2.9 million cycles (see Figure~\ref{fig:kyber_median_cycles_chart_log}). Average cycles were slightly higher, around 3.53 million, reflecting stable and repeatable enclave performance. Other key Kyber operations, such as \texttt{indcpa\_keypair} and \texttt{indcpa\_enc}, showed median cycle counts near 1.74 million and 2.62 million cycles respectively, consistent across runs.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=0.7\textwidth]{figures/kyber_64M_1Core_comparison.pdf}
%     \caption{Cycle count comparison for Kyber operations (native vs enclave). Error bars reflect variation over 10 runs.}
%     \label{fig:kyber_cycles}
% \end{figure}

@@

% \begin{figure}[h]
% \centering
% \begin{tikzpicture}

% % --- Left axis: Times + slowdown factors ---
% \begin{axis}[
%     ybar,
%     bar width=8pt,
%     width=14cm,
%     height=8cm,
%     ymin=0,
%     ymax=120,
%     ylabel={Time (s) / Slowdown ×},
%     symbolic x coords={Kyber,Dhrystone,CoreMark},
%     xtick=data,
%     enlarge x limits=0.35,
%     ymajorgrids,
%     grid style=dashed,
%     legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
%     axis y line*=left,
%     axis x line*=bottom,
%     bar shift=-6pt
% ]
% \addplot[pattern=north east lines,pattern color=black] 
%     coordinates {(Kyber,10.65) (Dhrystone,15.3) (CoreMark,101)};
% \addplot[pattern=dots,pattern color=black] 
%     coordinates {(Kyber,29.3) (Dhrystone,24.7) (CoreMark,107)};
% \addplot[pattern=crosshatch,pattern color=black] 
%     coordinates {(Kyber,2.6) (Dhrystone,1.6) (CoreMark,1.05)};
% \legend{Native time (s), Enclave time (s), Slowdown ×}
% \end{axis}

% % --- Right axis: Context switches ---
% \begin{axis}[
%     ybar,
%     bar width=8pt,
%     width=14cm,
%     height=8cm,
%     ymin=0,
%     ymax=70000,
%     ylabel={Involuntary Context Switches},
%     symbolic x coords={Kyber,Dhrystone,CoreMark},
%     xtick=data,
%     enlarge x limits=0.35,
%     hide x axis,
%     axis y line*=right,
%     axis x line=none,
%     bar shift=10pt
% ]
% \addplot[pattern=horizontal lines,pattern color=black] 
%     coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
% \legend{Involuntary CS}
% \end{axis}

% \end{tikzpicture}
% \caption{Native vs enclave performance: absolute times, slowdown factors (left axis), and involuntary context switches (right axis).}
% \end{figure}

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=12pt,
    width=12cm,
    height=7cm,
    ymin=0,
    ylabel={Value},
    symbolic x coords={RSS\ (KB), Minor\ PF, Major\ PF},
    xtick=data,
    enlarge x limits=0.25,
    ymajorgrids,
    grid style=dashed,
    legend style={at={(0.5,-0.18)},anchor=north,legend columns=-1}
]
% Native values
\addplot[pattern=north east lines,pattern color=black] coordinates
    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};
% Enclave values
\addplot[pattern=dots,pattern color=black] coordinates
    {(RSS\ (KB),10) (Minor\ PF,1) (Major\ PF,0)};

\legend{Native, Enclave}
\end{axis}
\end{tikzpicture}
\caption{Kyber benchmark: memory usage (RSS) and page fault counts for native vs enclave execution.}
\label{fig:memory_pagefaults_chart}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
% Left axis: CPU time percentages
\begin{axis}[
    ybar,
    bar width=8pt,
    width=14cm,
    height=8cm,
    ymin=0,
    ymax=100,
    ylabel={CPU Time (\%)},
    symbolic x coords={Kyber,Dhrystone,CoreMark},
    xtick=data,
    enlarge x limits=0.25,
    ymajorgrids,
    grid style=dashed,
    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
    axis y line*=left,
    axis x line*=bottom,
    nodes near coords,
    every node near coord/.append style={
        font=\small,
        anchor=west,
        rotate=90
    }
]

% User CPU% Native
\addplot[pattern=dots,pattern color=black] coordinates
    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};

% User CPU% Enclave
\addplot[pattern=dots,pattern color=black!50] coordinates
    {(Kyber,95) (Dhrystone,95) (CoreMark,95)};

% System CPU% Native
\addplot[pattern=crosshatch,pattern color=black] coordinates
    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};

% System CPU% Enclave
\addplot[pattern=crosshatch,pattern color=black!50] coordinates
    {(Kyber,5) (Dhrystone,5) (CoreMark,5)};

\legend{
    User CPU\% (Native),
    User CPU\% (Enclave),
    System CPU\% (Native),
    System CPU\% (Enclave),
    Ctx Switches (Native),
    Ctx Switches (Enclave)
}
\end{axis}

% Right axis: Context switches
\begin{axis}[
    ybar,
    bar width=8pt,
    width=14cm,
    height=8cm,
    ymin=0,
    ymax=70000,
    ylabel={Involuntary Context Switches},
    symbolic x coords={Kyber,Dhrystone,CoreMark},
    xtick=data,
    enlarge x limits=0.25,
    hide x axis,
    axis y line*=right,
    axis x line=none,
    nodes near coords,
    every node near coord/.append style={
        font=\small,
        anchor=west,
        rotate=90
    }
]

% Context switches Native
\addplot[pattern=north east lines,pattern color=black] coordinates
    {(Kyber,350) (Dhrystone,350) (CoreMark,350)};

% Context switches Enclave
\addplot[pattern=north east lines,pattern color=black!50] coordinates
    {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};

\end{axis}

\end{tikzpicture}
\caption{Native vs enclave execution: involuntary context switches (right axis) and CPU time percentages (left axis).}
\label{fig:context_switches_dualaxis}
\end{figure}

@@

In summary, enclave execution on this low-end RISC-V platform introduces moderate but consistent overheads, primarily in the form of increased runtime and context switching, without significant memory bloat or paging pressure. While sequential workloads complete reliably, attempts at concurrent enclave execution expose hard architectural limits rooted in PMP region constraints. These results highlight the current scalability boundaries of enclave systems and motivate the need for both software- and hardware-level support to unlock robust parallelism under secure execution.

\section{Comparative Overhead Metrics Across Workloads}
\label{sec:workload-sensitivity}

This section presents a comparative evaluation of the performance overheads introduced by secure enclave execution across three representative workloads: the Kyber post-quantum cryptographic scheme, the Dhrystone integer benchmark, and the CoreMark embedded systems benchmark. These workloads were selected for their diversity in computational characteristics—ranging from cryptography-heavy operations with irregular memory access patterns, to synthetic instruction-driven integer arithmetic, and finally to mixed workload execution involving control logic, arithmetic, and memory access. By comparing enclave and native performance across these distinct benchmarks, we aim to empirically characterize how enclave runtime behavior and hardware-level protections interact with different application classes.

All experiments were conducted on the same low-end RISC-V platform: a single in-order core with 64~MB of RAM and no hardware acceleration for encryption or virtualization. Both enclave and native executions used the same compiler flags, runtime configurations, and benchmarking scripts. Each benchmark was run ten times sequentially in both native and enclave modes to minimize measurement noise and provide stable median and average performance statistics. Measurements included CPU cycle counts, wall-clock time, throughput (where applicable), context switches, and memory footprint.

Among the workloads tested, Kyber was the most sensitive to enclave-related overhead. In native execution, the \texttt{kyber\_decaps} operation completed in a median of approximately 2.9 million cycles, corresponding to a wall-clock time of around 10.7 seconds. However, under enclave execution, the same operation required over 3.5 million cycles and a wall-clock time of 29.2 seconds, reflecting a 2.73$\times$ increase in observed duration. The increase in cycle count—approximately 20\%—was dwarfed by the nearly 3$\times$ increase in elapsed time, suggesting that the overhead is not merely computational but also systemic. The Kyber implementation is particularly sensitive to memory access latency due to its reliance on large data buffers, frequent pointer dereferencing, and memory-intensive operations such as polynomial multiplication and compression. These characteristics make it acutely vulnerable to the latency overhead introduced by enclave-specific features such as memory encryption and integrity verification, which intercept and transform nearly every memory read and write.

By contrast, CoreMark exhibited far greater resilience to enclave-induced slowdown. The native execution time averaged 101.5 seconds, while enclave execution required 107.3 seconds—a modest 1.06$\times$ slowdown. Throughput dropped slightly from a native range of 967--988 iterations per second to an enclave range of 919--939, indicating that while performance was affected, the impact was minimal relative to the workload's duration and structure. This result is consistent with CoreMark’s design as a balanced benchmark that includes control flow, integer operations, and limited memory activity. The longer total runtime may also help amortize fixed enclave setup and teardown costs over a larger instruction volume, dampening the perceived overhead.

Figure~\ref{fig:sensitivity-overheads-chart} summarizes the comparative performance and system-level behavior across all three benchmarks, illustrating the wall-clock time overheads and the uniformly high context switch counts observed during enclave execution.

Dhrystone, a classic synthetic benchmark designed to test CPU and compiler performance through tight integer loops and frequent procedure calls, showed intermediate sensitivity. Its native execution time averaged 15.3 seconds, while the enclave variant took 24.7 seconds—an overhead factor of approximately 1.61$\times$. Unlike Kyber, Dhrystone does not perform extensive memory operations, but its compact and time-sensitive control loops make it vulnerable to any systemic delay introduced by the enclave runtime, particularly if context switches or instruction fetch penalties are frequent. Its throughput dropped slightly, from around 400,000 operations per second natively to a range between 395,000 and 408,000 in the enclave, with the decline being small but consistent.

Across all three benchmarks, a consistent and striking overhead pattern emerged in the form of involuntary context switches. While native executions averaged fewer than 400 involuntary context switches per run, enclave executions across all workloads exhibited approximately 66,000 context switches per run—an increase of nearly two orders of magnitude. This effect was uniform across workloads, suggesting that it is driven by enclave runtime behavior rather than application-specific activity. Likely contributors include timer interrupts, enclave runtime transitions, secure fault handling, and scheduling-related activity under the Keystone framework. Despite this significant increase, the impact on user-mode computation was minimal in terms of CPU time distribution: more than 95\% of execution time was consistently spent in user space, indicating that the workload remained compute-bound and that the primary cost of context switching was time variability rather than computational interference.

Memory usage was consistent and tightly constrained across all runs. All workloads exhibited a resident set size (RSS) of approximately 10~KB in both enclave and native executions, with no major page faults and typically only a single minor page fault per run. This indicates that working sets for each benchmark fit entirely within the secure memory allocation provided by the enclave runtime, and that the paging subsystem—despite operating in a protected mode—remained effective and unobtrusive. This stability is particularly noteworthy given the limited RAM available and the additional overhead of enclave memory isolation mechanisms.

Taken together, the data suggests that the performance impact of enclave execution is highly dependent on the nature of the workload. Kyber, with its pointer-heavy, memory-intensive cryptographic routines, suffers the most under enclave constraints, particularly due to the latency introduced by secure memory mechanisms. Dhrystone, while less affected, still shows measurable sensitivity due to its reliance on fast, loop-driven control structures. CoreMark, in contrast, demonstrates a relatively minor slowdown, benefiting from workload diversity and longer runtime that dilutes per-access costs. At the system level, enclave-induced context switching is a universal phenomenon, independent of workload type, but it does not appear to disrupt user-mode execution beyond timing variability.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}

% --- Left axis: Execution times ---
\begin{axis}[
    ybar,
    bar width=6pt,
    width=14cm,
    height=8cm,
    ymin=0,
    ymax=120,
    ylabel={Execution Time (s)},
    symbolic x coords={Kyber,Dhrystone,CoreMark},
    xtick=data,
    enlarge x limits=0.4,
    ymajorgrids,
    grid style=dashed,
    legend style={at={(0.5,-0.18)},anchor=north,legend columns=3},
    axis y line*=left,
    axis x line*=bottom,
    bar shift=-4pt
]
% Native time
\addplot[pattern=dots,pattern color=black] 
    coordinates {(Kyber,10.7) (Dhrystone,15.3) (CoreMark,101.5)};

% Enclave time + overhead factor labels
\addplot[
    pattern=dots,
    pattern color=black!50,
    nodes near coords,
    point meta=explicit symbolic,
    every node near coord/.append style={
        font=\small,
        anchor=south,
        yshift=2pt,
        fill=white,
        fill opacity=0.7,
        text opacity=1
    }
] 
coordinates {
    (Kyber,29.2)      [2.73$\times$]
    (Dhrystone,24.7)  [1.61$\times$]
    (CoreMark,107.3)  [1.06$\times$]
};

\legend{Native Time, Enclave Time}
\end{axis}

% --- Right axis: Context switches ---
\begin{axis}[
    ybar,
    bar width=6pt,
    width=14cm,
    height=8cm,
    ymin=0,
    ymax=70000,
    ylabel={Involuntary Context Switches},
    symbolic x coords={Kyber,Dhrystone,CoreMark},
    xtick=data,
    enlarge x limits=0.4,
    hide x axis,
    axis y line*=right,
    axis x line=none,
    bar shift=10pt
]
\addplot[pattern=north east lines,pattern color=black] 
    coordinates {(Kyber,66000) (Dhrystone,66000) (CoreMark,66000)};
\legend{Involuntary CS}
\end{axis}

\end{tikzpicture}
\caption{Comparison of native and enclave execution times (left axis) with slowdown factors annotated, and involuntary context switch counts (right axis).}
\label{fig:sensitivity-overheads-chart}
\end{figure}
